Syntax and Semantics of Cedille

### 2023-10-23 docs/semantics/paper.tex
docs: semantics: some old unpushed changes
Signed-off-by: Christa Jenkins <cwjnkins@uiowa.edu>

\begin{theorem}[Syntactic kind preservation]
  \label{thm:syntactic-kind-pres}
  If \(\Gamma \vdash T \tpsynth \kappa\) and \(T \leadstoc T'\) then
  \(\Gamma \vdash T' \tpsynth \kappa'\) for some \(\kappa'\) such that \(\kappa
  \cong \kappa'\).
\end{theorem}

@@

Type checking \((\Gamma \vdash t \tpcheck T)\) is not covered in
Theorem~\ref{thm:judge-valid}, since the convention for a bidirectional system
is that there \(T\) is already \emph{assumed} to have type \(\star\) under a
typing context.

Lack of normalization for terms in general notwithstanding, it is still
important to confirm that the process of reducing types to weak head normal form
is itself terminating.
\begin{theorem}[Normalization of types under weak head normal form reduction]
  \label{thm:type-norm}
  \
  \begin{enumerate}
  \item If \(\Gamma \vdash \kappa\) then there exists \(\kappa'\) such that
    \(\kappa \leadsto^* \kappa' \nleadsto\)
  \end{enumerate} If \(\Gamma \vdash T : \kappa\) then there exists \(T'\) such
  that \(T \leadsto^* T' \nleadsto\).
\end{theorem}

@@

\begin{corollary}
  \ \\
  \begin{itemize}
  \item If \(\vdash \Gamma_1,x:T,\Gamma_2\) and \(\Gamma_1 \vdash t \tpsynth T\)
    then \(\vdash \Gamma_1,[t/x]\Gamma_2\)

@@

\subsection{Theorem~\ref{thm:syntactic-kind-pres}}
\begin{proof}[Proof (of Theorem~\ref{thm:syntactic-kind-pres})]
  We may rule out the cases where \(T\) is a variable or formed by a type
  constructor, as this would contradict the assumption that \(T \leadstoc
  T'\) for some \(T'\).
  We omit type annotations from substitutions when they are clear from the context.

  \startcase{.2cm}
  \[
    \infer{\Gamma\vdash T\ t \tpsynth [t/x]^{T_1}\kappa_1}{\Gamma\vdash T \tpsynth
      \abs{\Pi}{x}{T_1}{\kappa_1} & \Gamma\vdash t \tpcheck T_1}
  \]

  There are two subcases to consider for the derivation of \(T\ t \leadstoc
  T'\). In the first case, we have \(T \leadstoc T''\) for some \(T''\) (so
  \(T' = T''\ t\)). By the IH, we have \(\Gamma \vdash T'' \tpsynth
  \abs{\Pi}{x}{T_1'}{\kappa_1'}\) for some \(T_1' \cong T_1\) and \(\kappa_1'
  \cong \kappa_1\) (we have from that IH that kind of \(T''\) must be
  convertible with \(\abs{\Pi}{x}{T_1}{\kappa_1}\)). By
  Lemma~\ref{lem:ctxt-conv-class2}, \(\Gamma \vdash t \tpcheck T_1'\).
  By the rule, \(\Gamma \vdash T''\ t \tpsynth [t/x]\kappa_1'\), and the
  synthesized kind is clearly convertible with \([t/x]\kappa_1\).

  In the other subcase, the subject of kinding is of the form
  \((\abs{\lambda}{x}{T_1}{T''})\ t\) and our assumed reduction is to
  \([t/x]T''\). By inversion of the kinding derivation, we have \(\Gamma,x:T_1
  \vdash T'' \tpsynth \kappa_1\). By Lemma~\ref{lem:subst-class}, we have
  \(\Gamma \vdash [t/x]T'' \tpsynth [t/x]\kappa_1\).

  \startcase{.2cm}
  \[
    \infer{\Gamma\vdash T_1 \cdot T_2 \tpsynth [T_2/X]\kappa_1}{\Gamma\vdash T_1
      \tpsynth \abs{\Pi}{X}{\kappa_2}{\kappa_1} & \Gamma\vdash T_2 \tpsynth
      \kappa_2' & \kappa_2 \cong \kappa_2'}
  \]

  There are two subcases to consider for the derivation of \(T_1 \cdot T_2
  \leadstoc T'\). In the first subcase, we have \(T_1 \leadstoc
  T_1''\) for some \(T_1''\) (so \(T' = T_1'' \cdot T_2\)). By the IH, we have
  \(\Gamma \vdash T_1'' \tpsynth \abs{\Pi}{X}{\kappa_2''}{\kappa_1''}\) for some
  \(\kappa_2'' \cong \kappa_2\) and \(\kappa_1'' \cong \kappa_1\) (we have from
  the IH that the kind of \(T_1''\) must be convertible with
  \(\abs{\Pi}{X}{\kappa_2}{\kappa_1}\)).
  By transitivity, we can conclude \(\kappa_2'' \cong \kappa_2'\).
  By the rule, we have \(\Gamma \vdash T_1'' \cdot T_2 \tpsynth
  [T_2/X]\kappa_1''\), and the synthesized kind is clearly convertible with
  \([T_2/X]\kappa_1\).

  In the other subcase, the subject of kinding is of the form
  \((\abs{\lambda}{X}{\kappa_2}{T_1''}) \cdot T_2\) and our assumed reduction is
  to \([T_2/X]T_1''\).
  By inversion of the kinding derivation, we have \(\Gamma,X:\kappa_2 \vdash
  T_1'' \tpsynth \kappa_1\).
  By Lemma~\ref{lem:ctxt-conv-class}, we have \(\Gamma,X:\kappa_2' \vdash T_1''
  \tpsynth \kappa_1'\) for some \(\kappa_1' \cong \kappa_1\), and from this and
  Lemma~\ref{lem:subst-class} we have \(\Gamma \vdash [T_2/X]T_1'' \tpsynth
  [T_2/X]\kappa_1'\).
  This is clearly convertible with \([T_2/X]\kappa_1\), as desired.
\end{proof}

@@

Straightaway, we see that, when translated as types, all CDLE kinds (well-formed
or not) are translated to well-formed types of this calculus, and
well-formedness of contexts is preserved by this translation.
It is also not hard to see that convertible kinds are translated to equal types,
and that substitution in kinds is a no-op.
\begin{proposition}
  \label{thm:stlc-kinds}
  \ \\
  \begin{enumerate}
  \item \(\interp{\kappa}_{\text{K}}\ \mathit{wf}\) for all \(\kappa\)

@@

  \item If \(T \leadsto T'\) then \(\interp{T} \leadsto^+ \interp{T'}\)
  \end{enumerate}
\end{lemma}
\begin{proof}
  By mutual induction on the derivations of \(\kappa \leadsto \kappa'\) and \(T
  \leadsto T'\).
  We implicitly make use of the property that if \(T \leadsto^+ T'\) (or
  \(\kappa \leadsto^+ \kappa'\)), then this reduction can be lifted to any
  contextual closure of the classifier.

@@

  By the IH, \(\interp{\kappa} \leadsto \interp{\kappa'}\).
  Deduce
  \[
    \infer{
      \mathit{con}(X\!:\!\interp{\kappa}_{\text{K}},\interp{\kappa},\interp{T})
      \leadsto \mathit{con}(X\!:\!\interp{\kappa}_{\text{K}},\interp{\kappa'},\interp{T})
    }{
      \interp{\kappa} \leadsto \interp{\kappa'}
    }
  \]

@@

We can now that reduction of CDLE types and kinds, minus reduction of terms, is
strongly normalizing.
\begin{proof}[Proof (of Theorem~\ref{thm:type-norm})]
  By contradiction. Assume we have an infinite reduction sequence starting from
  \(T\).
  By Lemma~\ref{thm:stlc-static-sound}, \(\interp{\Gamma} \vdash \interp{T} :
  \interp{\kappa}\), and by coiteratively applying
  Lemma~\ref{thm:stlc-dynamic-sound} we have constructed an infinite reduction
  sequence starting from the well-typed STLC term \(\interp{T}\), contradicting
  strong normalization of the calculus.

>>>>

\begin{theorem}[Syntactic kind preservation]
  \label{thm:syntactic-kind-pres}
  \
  \begin{itemize}
  \item If \(\Gamma \vdash \kappa\) and \(\kappa \rightsquigarrow \kappa'\) then
    \(\Gamma \vdash \kappa'\)

  \item If \(\Gamma \vdash T \tpsynth \kappa\) and \(T \leadstoc T'\) then
    \(\Gamma \vdash T' \tpsynth \kappa'\) for some \(\kappa'\) such that \(\kappa
    \cong \kappa'\).
  \end{itemize}

\end{theorem}

@@

Type checking \((\Gamma \vdash t \tpcheck T)\) is not covered in
Theorem~\ref{thm:judge-valid}, since the convention for a bidirectional system
is that there \(T\) is already \emph{assumed} to have type \(\star\) under the
given typing context.

Lack of normalization for terms in general notwithstanding, it is still
important to confirm that the process of reducing types to weak head normal form
is itself terminating.
\begin{theorem}[Normalization of types and kinds under term-free reduction]
  \label{thm:type-norm}
  \
  \begin{enumerate}
  \item If \(\Gamma \vdash \kappa\) then there exists \(\kappa'\) such that
    \(\kappa \leadsto^* \kappa' \nleadsto\)
   \item If \(\Gamma \vdash T : \kappa\) then there exists \(T'\) such
     that \(T \leadsto^* T' \nleadsto\).
 \end{enumerate}
\end{theorem}

@@

\begin{corollary}
  \begin{itemize}
  \item If \(\vdash \Gamma_1,x:T,\Gamma_2\) and \(\Gamma_1 \vdash t \tpsynth T\)
    then \(\vdash \Gamma_1,[t/x]\Gamma_2\)

@@

\subsection{Theorem~\ref{thm:syntactic-kind-pres}}
\begin{proof}[Proof (of Theorem~\ref{thm:syntactic-kind-pres})]
  By mutual induction on the derivation of \(\kappa \rightsquigarrow \kappa'\)
  or \(T \rightsquigarrow T'\), depending on the case.
  We omit type annotations from substitutions when they are clear from the context.

  \startcase{.2cm}
  \[
    \infer{
      (\abs{\lambda}{x}{T_1}{T_2})\ t \rightsquigarrow T_2[t/x]
    }{}
  \]

  By inversion of assumed the typing derivation,
  \[
    \infer{
      \Gamma \vdash (\abs{\lambda}{x}{T_1}{T_2})\ t \tpsynth \kappa[t/x]
    }{
      \infer{
        \Gamma \vdash \abs{\lambda}{x}{T_1}{T_2} \tpsynth \abs{\Pi}{x}{T_1}{\kappa}
      }{
        \Gamma,x\!:\!T_1 \vdash T_2 \tpsynth \kappa
      }
      &
      \Gamma \vdash t \tpcheck T_1
    }
  \]

  By Lemma~\ref{lem:subst-class}, \(\Gamma \vdash T_2[t/x] \tpsynth
  \kappa[t/x]\), as desired.

  \startcase{.2cm}
  \[
    \infer{
      (\abs{\lambda}{X}{\kappa}T) \cdot T' \rightsquigarrow T[T'/X]
    }{}
  \]

  Similar to the above case

  \startcase{.2cm}
  \[
    \infer{
      T_1\ t_2 \rightsquigarrow T_1'\ t_2
    }{
      T_1 \rightsquigarrow T_1'
    }
  \]

  By inversion of the assumed typing derivation,
  \[
    \infer{
      \Gamma \vdash T_1\ t_2 \tpsynth \kappa[t/x]
    }{
      \Gamma \vdash T_1 \tpsynth \abs{\Pi}{x}{T_2}{\kappa}
      & \Gamma \vdash t_2 \tpcheck T_2
    }
  \]

  By the IH and inversion of kind conversion, \(\Gamma \vdash T_1' \tpsynth
  \abs{\Pi}{x}{T_2'}{\kappa'}\) with \(T_2 \cong T_2'\) and \(\kappa \cong
  \kappa'\).
  As it is clear that \(\kappa \cong \kappa'\) implies \(\kappa[t_2/x]
  \cong \kappa[t_2/x]\), and also by Lemma~\ref{lem:ctxt-conv-class2}   that
  \(\Gamma \vdash t_2 \tpcheck T_2\) implies \(\Gamma \vdash t_2 \tpcheck
  T_2'\), we deduce
  % TODO
  \[
    \infer{
      \Gamma \vdash T_1'\ t_2 \tpsynth \kappa'[t_2/x]
    }{
      \Gamma \vdash T_1 \tpsynth \abs{\Pi}{x}{T_2'}{\kappa'}
      \quad \Gamma \vdash t_2 \tpcheck T_2
    }
  \]

  \startcase{.2cm}
  \[
    \infer{
      T_1 \cdot T_2 \rightsquigarrow T_1' \cdot T_2
    }{
      T_1 \rightsquigarrow T_1'
    }
  \]

  Similar to the above case, using transitivity of \(\cong\) for kinds.

  \startcase{.2cm}
  \[
    \infer{
      T_1 \cdot T_2 \rightsquigarrow T_1 \cdot T_2'
    }{
      T_2 \rightsquigarrow T_2'
    }
  \]

  By inversion of the typing derivation,
  \[
    \infer{
      \Gamma \vdash T_1 \cdot T_2 \tpsynth \kappa_1[T_2/X]
    }{
      \Gamma \vdash T_1 \tpsynth \abs{\Pi}{X}{\kappa_2}{\kappa_1}
      & \Gamma \vdash T_2 \tpsynth \kappa_2'
      & \kappa_2 \cong \kappa_2'
    }
  \]

  By the IH, \(\Gamma \vdash T_2 \tpsynth \kappa_2''\) for some \(\kappa_2''
  \cong \kappa_2'\).
  By transitivity of \(\cong\), \(\kappa_2 \cong \kappa_2''\).
  Deduce
  \[
    \infer{
      \Gamma \vdash T_1 \cdot T_2' \tpsynth \kappa_1[T_2'/X]
    }{
      \Gamma \vdash T_1 \tpsynth \abs{\Pi}{X}{\kappa_2}{\kappa_1}
      & \Gamma \vdash T_2 \tpsynth \kappa_2''
      & \kappa_2 \cong \kappa_2''
    }
  \]

  \startcase{.2cm}
  \[
    \infer{
      \abs{\mathcal{Q}}{x}{T_1}{T_2} \rightsquigarrow \abs{\mathcal{Q}}{x}{T_1'}{T_2}
    }{
      T_1 \rightsquigarrow T_1'
      & \mathcal{Q} \in \{\Pi,\forall,\iota\}
    }
  \]

  By inversion of the kinding judgment,
  \[
    \infer{
      \Gamma \vdash \abs{\mathcal{Q}}{x}{T_1}{T_2} \tpsynth \star
    }{
      \Gamma \vdash T_1 \tpsynth \star
      & \Gamma,x\!:\!T_1 \vdash T_2 \tpsynth \star
    }
  \]

  By the IH and inversion of kind conversion, \(\Gamma \vdash T_1' \tpsynth
  \star\).
  By Lemma~\ref{lem:ctxt-conv-class2}, \(\Gamma,x\!:\!T_1' \vdash T_2 \tpsynth
  \star\) (\(T_1 \rightsquigarrow T_1'\) implies \(T_1 \cong T_1'\)).
  Deduce
  \[
    \infer{
      \Gamma \vdash \abs{\mathcal{Q}}{x}{T_1'}{T_2} \tpsynth \star
    }{
      \Gamma \vdash T_1' \tpsynth \star
      & \Gamma, x\!:\!T_1' \vdash T_2 \tpsynth \star
      & \mathcal{Q} \in \{\Pi,\forall,\iota\}
    }
  \]

  \startcase{.2cm}
  \[
    \infer{
      \abs{\forall}{X}{\kappa}{T} \rightsquigarrow \abs{\forall}{X}{\kappa'}{T}
    }{
      \kappa \rightsquigarrow \kappa'
    }
  \]

  By inversion of the kinding judgment,
  \[
    \infer{
      \Gamma \vdash \abs{\forall}{X}{\kappa}{T} \tpsynth \star
    }{
      \Gamma \vdash \kappa
      & \Gamma,X\!:\!\kappa \vdash T \tpsynth \star
    }
  \]

  By the IH, \(\Gamma \vdash \kappa'\).
  By Lemma~\ref{lem:ctxt-conv-class}, \(\Gamma,X\!:\kappa' \vdash T \tpsynth \star\).
  Deduce
  \[
    \infer{
      \Gamma \vdash \abs{\forall}{X}{\kappa'}{T} \tpsynth \star
    }{
      \Gamma \vdash \kappa'
      & \Gamma,X\!:\!\kappa' \vdash T \tpsynth \star
    }
  \]

  \startcase{.2cm}
  \[
    \infer{
      \abs{\mathcal{Q}}{x}{T_1}{T_2}
      \rightsquigarrow
      \abs{\mathcal{Q}}{x}{T_1}{T_2'}
    }{
      T_2 \rightsquigarrow T_2'
    }
  \]

  By inversion of the kinding rule,
  \[
    \infer{
      \Gamma \vdash \abs{\mathcal{Q}}{x}{T_1}{T_2} \tpsynth \star
    }{
      \Gamma \vdash T_1 \tpsynth \star
      & \Gamma,x\!:\!T_1 \vdash T_2 \tpsynth \star
    }
  \]

  By the IH and inversion of kind conversion, \(\Gamma,x\!:\!T_1 \vdash T_2'
  \tpsynth \star\).
  Deduce
  \[
    \infer{
      \Gamma \vdash \abs{\mathcal{Q}}{x}{T_1}{T_2'} \tpsynth \star
    }{
      \Gamma \vdash T_1 \tpsynth \star
      & \Gamma,x\!:\!T_1 \vdash T_2' \tpsynth \star
    }
  \]

  \startcase{.2cm}
  \[
    \infer{
      \abs{\forall}{X}{\kappa}{T} \rightsquigarrow \abs{\forall}{X}{\kappa}{T'}
    }{
      T \rightsquigarrow T'
    }
  \]

  \startcase{.2cm}
  \[
    \infer{
      \abs{\Pi}{x}{T}{\kappa} \rightsquigarrow \abs{\Pi}{x}{T'}{\kappa'}
    }{
      T \rightsquigarrow T'
    }
  \]

  By inversion of the kinding derivation,
  \[
    \infer{
      \Gamma \vdash \abs{\Pi}{x}{T}{\kappa}
    }{
      \Gamma \vdash T \tpsynth \star
      \quad \Gamma,x\!:\!T \vdash \kappa
    }
  \]

  By the IH, \(\Gamma \vdash T' \tpsynth \star\).
  By Lemma~\ref{lem:ctxt-conv-class2}, \(\Gamma,x\!:\!T' \vdash \kappa\).
  Deduce
  \[
    \infer{
      \Gamma \vdash \abs{\Pi}{x}{T'}{\kappa}
    }{
      \Gamma \vdash T' \tpsynth \star
      \quad \Gamma,x\!:\!T' \vdash \kappa
    }
  \]

  \startcase{.2cm}
  \[
    \infer{
      \abs{\Pi}{X}{\kappa_1}{\kappa_2} \rightsquigarrow \abs{\Pi}{X}{\kappa_1'}{\kappa_2}
    }{
      \kappa_1 \rightsquigarrow \kappa_1'
    }
  \]

  Similar to the above case, appealing to Lemma~\ref{lem:ctxt-conv-class}.

  \startcase{.2cm}
  \[
    \infer{
      \abs{\Pi}{x}{T}{\kappa} \rightsquigarrow \abs{\Pi}{x}{T}{\kappa'}
    }{
      \kappa \rightsquigarrow \kappa'
    }
  \]

  By inversion of the kinding derivation,
  \[
    \infer{
      \Gamma \vdash \abs{\Pi}{x}{T}{\kappa}
    }{
      \Gamma \vdash T \tpsynth \star
      \quad \Gamma,x\!:\!T \vdash \kappa
    }
  \]

  By the IH, \(\Gamma,x\!:\!T \vdash \kappa'\).
  Deduce
  \[
    \infer{
      \Gamma \vdash \abs{\Pi}{x}{T}{\kappa'}
    }{
      \Gamma \vdash T \tpsynth \star
      \quad \Gamma,x\!:\!T \vdash \kappa'
    }
  \]

  \startcase{.2cm}
  \[
    \infer{
      \abs{\Pi}{X}{\kappa_1}{\kappa_2} \rightsquigarrow \abs{\Pi}{X}{\kappa_1}{\kappa_2'}
    }{
      \kappa_2 \rightsquigarrow \kappa_2'
    }
  \]

@@

Straightaway, we see that, when translated as types, all CDLE kinds (well-formed
or not) are translated to well-formed types of this calculus, and
well-formedness of contexts is preserved by this translation.
It is also not hard to see that convertible kinds are translated to equal types,
and that substitution in kinds is a no-op.
\begin{proposition}
  \label{thm:stlc-kinds}
  \ 
  \begin{enumerate}
  \item \(\interp{\kappa}_{\text{K}}\ \mathit{wf}\) for all \(\kappa\)

@@

  \item If \(T \leadsto T'\) then \(\interp{T} \leadsto^+ \interp{T'}\)
  \end{enumerate}

  Where \(\rightsquigarrow^+ =_{\textbf{df}}\ (\rightsquigarrow)\ ;\ (\rightsquigarrow^*)\)
\end{lemma}
\begin{proof}
  By mutual induction on the derivations of \(\kappa \leadsto \kappa'\) and \(T
  \leadsto T'\).
  We implicitly make use of the property that if \(T \leadsto^+ T'\) (or
  \(\kappa \leadsto^+ \kappa'\)), then this reduction can be lifted to any
  contextual closure of the classifier.

@@

  By the IH, \(\interp{\kappa} \leadsto \interp{\kappa'}\).
  Deduce
  \[
    \infer{
      \mathit{con}(X\!:\!\interp{\kappa}_{\text{K}},\interp{\kappa},\interp{T})
      \leadsto^+
      \mathit{con}(X\!:\!\interp{\kappa}_{\text{K}},\interp{\kappa'},\interp{T})
    }{
      \interp{\kappa} \leadsto^+ \interp{\kappa'}
    }
  \]

@@

We can now that reduction of CDLE types and kinds, minus reduction of terms, is
strongly normalizing.
\begin{proof}[Proof (of Theorem~\ref{thm:type-norm})]
  By contradiction. Assume we have an infinite reduction sequence starting from
  \(T\).
  By Lemma~\ref{thm:stlc-static-sound}, \(\interp{\Gamma} \vdash \interp{T} :
  \interp{\kappa}_{\text{K}}\), and by coiteratively applying
  Lemma~\ref{thm:stlc-dynamic-sound} we have constructed an infinite reduction
  sequence starting from the well-typed STLC term \(\interp{T}\), contradicting
  strong normalization of the calculus.
